---
title: "CodeBook"
author: "JimSwainston"
date: "August 23, 2015"
output: html_document
---

## Project Description
The goal of this project was to prepare a tidy data set with data collected from the accelerometers from the Samsung Galaxy S smartphone. 
The steps we were asked to complete were:

   Create one R script called run_analysis.R that does the following: 

   1.  Merges the training and the test sets to create one data set.
   2.  Extracts only the measurements on the mean and standard deviation for each measurement. 
   3.  Uses descriptive activity names to name the activities in the data set
   4.  Appropriately labels the data set with descriptive variable names. 
   5.  From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject.


###Collection of the raw data
The raw data set was downloaded from https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip

####Notes on the original raw data 

The readme file from the original dataset provides the following details of how the data was collected:

The experiments have been carried out with a group of 30 volunteers within an age bracket of 19-48 years. Each person performed six activities (WALKING, WALKING_UPSTAIRS, WALKING_DOWNSTAIRS, SITTING, STANDING, LAYING) wearing a smartphone (Samsung Galaxy S II) on the waist. Using its embedded accelerometer and gyroscope, we captured 3-axial linear acceleration and 3-axial angular velocity at a constant rate of 50Hz. The experiments have been video-recorded to label the data manually. The obtained dataset has been randomly partitioned into two sets, where 70% of the volunteers was selected for generating the training data and 30% the test data. 

The sensor signals (accelerometer and gyroscope) were pre-processed by applying noise filters and then sampled in fixed-width sliding windows of 2.56 sec and 50% overlap (128 readings/window). The sensor acceleration signal, which has gravitational and body motion components, was separated using a Butterworth low-pass filter into body acceleration and gravity. The gravitational force is assumed to have only low frequency components, therefore a filter with 0.3 Hz cutoff frequency was used. From each window, a vector of features was obtained by calculating variables from the time and frequency domain. See 'features_info.txt' for more details. 

For each record it is provided:

- Triaxial acceleration from the accelerometer (total acceleration) and the estimated body acceleration.
- Triaxial Angular velocity from the gyroscope. 
- A 561-feature vector with time and frequency domain variables. 
- Its activity label. 
- An identifier of the subject who carried out the experiment.


##Creating the tidy datafile
The run_analysis.r script has been written to output a tidy data file that adheres to the following principles of tidy data:

* Each variable forms a column
* Each observation forms a row
* Each type of observational unit forms a table

The data file gives averages of a number of variables for each activity and each subject.

###Guide to create the tidy data file
1. Download the raw data from from https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip

2. Copy the following data files to your working directory 
  + X_test.txt - The test data set
  + features.txt - variable names for the training and test data
  + y_test.txt - activity identifiers for the test set
  + subject_test.txt - subject indentifiers for the test set 
  + X_train.txt - The training data set
  + y_train.txt - activity identifiers for the training set
  + subject_train.txt - subject indentifiers for the training set 
  
3. Load the funciton run_analysis() that is contained in the script run_analysis.r. Note: this is dependent on the package dplyr.
  + Load the function 
  ```{r}
   source('run_analysis.r')
  ```
  
4. Run the function run_analysis()
  + check that the file tidy_data.txt has been created in your working directory
  
5. View the tidy data file

```{r}
data <- read.table("tidy_data.txt", header = TRUE)
View(data)
```


###Cleaning of the data
1. The cleaning script first renames all of the column names in the test and training datasets with the variable names in features.txt
2. The activity code variables in the y_train and y_test files are replaced with their meaningful data (e.g. 1 replaced with WALKGING) and then bound the respective train and test data sets. Column name renamed as activity_label.
3. The subect identifier variables in the subject_train and subject_test files are bound to their respective train and test data sets. Column name renamed as subject_id.
4. The training data and test data is stacked to create one full table containing both. 
5. The table is reduced so it only contains columns that contain mean() or std() in the variable name plus the activity and subject identifiers
6. The average of each variable is calculated for each activity and subject.
7. Columns renamed to reflect the data is now an average
8. Columns renamed to give better variable description. E.g. Acc replaced with Accelerometer for accelerometer measurements
9. The tidy data file is written to a txt file. 


##Description of the variables in the tidy_data.txt file

General description of the file including:

 - Dimensions of the dataset
 
[1] 180  68
 
 - Summary of the data
 
 Average of each mean and standard deviation measurement for each activity and subject using data collected from the accelerometers from the Samsung Galaxy S smartphone
 
##Variables present in the dataset
 
###subject_id 
integer identifier of the subject who performed the activity. Its range is from 1 to 30.

###activity_label
String indentifier of the activity performed. One of 6 fixed types:

* WALKING
* WALKING_UPSTAIRS
* WALKING_DOWNSTAIRS
* SITTING
* STANDING
* LAYING

###THE BELOW VARIABLES ARE DESCRIBED AS A GROUP AS THEY FOLLOW THE SAME SCHEMA

numeric data type ranging between 1 and -1

variable naming scheme:

Each variable is recorded along three axis: x, y & z

* time indicates a time domain measurement
* frequency indicates a frequency domain measurement
* accelerometer indicates an acceleration measurement from the accelerometer sensor
* gyroscope indicates an angular velocity measurement from the gyroscope sensor
* body indicates a body acceleration signal
* gravity indicates a gravity acceleration signal
* jerk indicates a jerk signal
* magnitude indicates a magnitude reading of the signal

all variables are averages for each subject and activity

###timeBodyAccelerometer-mean()-X-average 
###timeBodyAccelerometer-mean()-Y-average 
###timeBodyAccelerometer-mean()-Z-average 
###timeBodyAccelerometer-std()-X-average 
###timeBodyAccelerometer-std()-Y-average 
###timeBodyAccelerometer-std()-Z-average 
###timeGravityAccelerometer-mean()-X-average 
###timeGravityAccelerometer-mean()-Y-average 
###timeGravityAccelerometer-mean()-Z-average 
###timeGravityAccelerometer-std()-X-average 
###timeGravityAccelerometer-std()-Y-average 
###timeGravityAccelerometer-std()-Z-average 
###timeBodyAccelerometerJerk-mean()-X-average 
###timeBodyAccelerometerJerk-mean()-Y-average 
###timeBodyAccelerometerJerk-mean()-Z-average 
###timeBodyAccelerometerJerk-std()-X-average 
###timeBodyAccelerometerJerk-std()-Y-average 
###timeBodyAccelerometerJerk-std()-Z-average 
###timeBodyGyroscope-mean()-X-average 
###timeBodyGyroscope-mean()-Y-average 
###timeBodyGyroscope-mean()-Z-average 
###timeBodyGyroscope-std()-X-average 
###timeBodyGyroscope-std()-Y-average 
###timeBodyGyroscope-std()-Z-average 
###timeBodyGyroscopeJerk-mean()-X-average 
###timeBodyGyroscopeJerk-mean()-Y-average 
###timeBodyGyroscopeJerk-mean()-Z-average 
###timeBodyGyroscopeJerk-std()-X-average 
###timeBodyGyroscopeJerk-std()-Y-average 
###timeBodyGyroscopeJerk-std()-Z-average 
###timeBodyAccelerometerMagnitude-mean()-average 
###timeBodyAccelerometerMagnitude-std()-average 
###timeGravityAccelerometerMagnitude-mean()-average 
###timeGravityAccelerometerMagnitude-std()-average 
###timeBodyAccelerometerJerkMagnitude-mean()-average 
###timeBodyAccelerometerJerkMagnitude-std()-average 
###timeBodyGyroscopeMagnitude-mean()-average 
###timeBodyGyroscopeMagnitude-std()-average 
###timeBodyGyroscopeJerkMagnitude-mean()-average 
###timeBodyGyroscopeJerkMagnitude-std()-average 
###frequencyBodyAccelerometer-mean()-X-average 
###frequencyBodyAccelerometer-mean()-Y-average 
###frequencyBodyAccelerometer-mean()-Z-average 
###frequencyBodyAccelerometer-std()-X-average 
###frequencyBodyAccelerometer-std()-Y-average 
###frequencyBodyAccelerometer-std()-Z-average 
###frequencyBodyAccelerometerJerk-mean()-X-average 
###frequencyBodyAccelerometerJerk-mean()-Y-average 
###frequencyBodyAccelerometerJerk-mean()-Z-average 
###frequencyBodyAccelerometerJerk-std()-X-average 
###frequencyBodyAccelerometerJerk-std()-Y-average 
###frequencyBodyAccelerometerJerk-std()-Z-average 
###frequencyBodyGyroscope-mean()-X-average 
###frequencyBodyGyroscope-mean()-Y-average 
###frequencyBodyGyroscope-mean()-Z-average 
###frequencyBodyGyroscope-std()-X-average 
###frequencyBodyGyroscope-std()-Y-average 
###frequencyBodyGyroscope-std()-Z-average 
###frequencyBodyAccelerometerMagnitude-mean()-average 
###frequencyBodyAccelerometerMagnitude-std()-average 
###frequencyBodyBodyAccelerometerJerkMagnitude-mean()-average 
###frequencyBodyBodyAccelerometerJerkMagnitude-std()-average 
###frequencyBodyBodyGyroscopeMagnitude-mean()-average 
###frequencyBodyBodyGyroscopeMagnitude-std()-average 
###frequencyBodyBodyGyroscopeJerkMagnitude-mean()-average 
###frequencyBodyBodyGyroscopeJerkMagnitude-std()-average