---
title: "README"
author: "JimSwainston"
date: "August 23, 2015"
output: html_document
---

## Script Description

The run_analysis.r script has been written to output a tidy data file as per the project specification. (see codebook)

### What does the script do?

The script has been written to output a tidy data file that adheres to the following principles of tidy data:

* Each variable forms a column
* Each observation forms a row
* Each type of observational unit forms a table

The following steps give an overiew of how the script cleans the data and outputs a tidy data file.

1. The cleaning script first renames all of the column names in the test and training datasets with the variable names in features.txt
2. The activity code variables in the y_train and y_test files are replaced with their meaningful data (e.g. 1 replaced with WALKGING) and then bound the respective train and test data sets. Column name renamed as activity_label.
3. The subect identifier variables in the subject_train and subject_test files are bound to their respective train and test data sets. Column name renamed as subject_id.
4. The training data and test data is stacked to create one full table containing both. 
5. The table is reduced so it only contains columns that contain mean() or std() in the variable name plus the activity and subject identifiers
6. The average of each variable is calculated for each activity and subject.
7. Columns renamed to reflect the data is now an average
8. Columns renamed to give better variable description. E.g. Acc replaced with Accelerometer for accelerometer measurements
9. The tidy data file is written to a txt file. 

The data file gives averages of a number of variables for each activity and each subject.

### How do I run the script and view the results?

1. Download the raw data from from https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip

2. Copy the following data files to your working directory 
  + X_test.txt - The test data set
  + features.txt - variable names for the training and test data
  + y_test.txt - activity identifiers for the test set
  + subject_test.txt - subject indentifiers for the test set 
  + X_train.txt - The training data set
  + y_train.txt - activity identifiers for the training set
  + subject_train.txt - subject indentifiers for the training set 
  
3. Load and the funciton run_analysis() that is contained in the script run_analysis.r. Note: this is dependent on the package dplyr.
  + Load the script and run function
```{r,eval=FALSE}
   source('run_analysis.r')
```
  
```{r,eval=FALSE}
   run_analysis()
```
4. Check that the file tidy_data.txt has been created in your working directory
  
5. View the tidy data file

```{r,eval=FALSE}
data <- read.table("tidy_data.txt", header = TRUE)
View(data)
```


